---
title: "metabolomicsProcessingDIY"
author: 
- "DeniseSl22"
- "YourGithubUserName"
date: "17/05/24"
output:
 md_document:
    variant: markdown_github
always_allow_html: true
---
## Introduction
In this section of the workflow, we will obtain the metabolomics data and apply filtering options, to create a dataset ready for further statistical and pathway analysis (on Thursday).

## First, we setup the required libraries to get started.
```{r setup, warning=FALSE, message=FALSE}
# check if libraries are already installed > otherwise install it
if(!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager",repos = "http://cran.us.r-project.org")
if(!"dplyr" %in% installed.packages()) BiocManager::install("dplyr")
if(!"stringr" %in% installed.packages()){install.packages("stringr")}
#load libraries
library(dplyr)
library(stringr)
```

## Locate the data (in the file explorer) and load it in R
```{r}
if(!"readxl" %in% installed.packages()){install.packages("readxl")}
library("readxl")

dataLocation <- paste0(getwd(), '/Data/NoMa_NMR_intervention_GEOcodes_Sent Susan.xlsx')
metabolomicsData <- read_excel(dataLocation, 1)  # for first tab

#Make a copy of the data to compare against if needed
metabolomicsDataCopy <- metabolomicsData
```

## Second programming tasks: check NAs
```{r}
#Use the search box to find sample ‘2_A5’

# Convert the textual 'NA' labels with the real NA (not available)
metabolomicsData[metabolomicsData == "NA"] <- NA

# Check how many NAs are part of sample '2_A5'.
sum(is.na(metabolomicsData$`Sample name` == "2_A5")) ##Answer is 98

# Install the required package
if(!"heatmaply" %in% installed.packages()){install.packages("heatmaply")}
library("heatmaply")

# Visualize this information in a heatmap
heatmaply::heatmaply(is.na10(metabolomicsData[,3:100], grid_gap = 1, colors = heat.colors(200), showticklabels = c(T, F), margins = c(80, 10)))
```

## Third programming tasks check data correlations
```{r}
# Make sure all data is captured as numbers (numeric!)
metabolomicsData <- data.frame(metabolomicsData[,1:2], lapply(metabolomicsData[,3:length(metabolomicsData)],as.numeric))

# Remove rows with too many NAs
metabolomicsData <- metabolomicsData[complete.cases(metabolomicsData[ , 3:length(metabolomicsData)]),]

# Visualize all information in a heatmap
heatmaply::heatmaply(metabolomicsData[,3:100], grid_gap = 0, colors = heat.colors(200), showticklabels = c(T, F), margins = c(80, 10))
```

## Fourth programming tasks: check data distribution
```{r}
# Filter out data if the 'Sample name' column is NA 
metabolomicsData <- subset(metabolomicsData, !is.na(metabolomicsData$Sample.name))

# Inspect the remaining samples names; do any stand out (in a negative way)? If yes, remove these
metabolomicsData <- subset(metabolomicsData, metabolomicsData$Sample.name != 'missing Gex')

# Download the file 'targets.csv' from Surfdrive, add this to the Data folder, and change the location of data you want to load
dataLocation <- paste0(getwd(), '/Data/targets.csv')

# Load the targets data
treatmentData <- read.csv(dataLocation, header = TRUE)

# Clean up the sample names (since they are not the same for metabolomics)
treatmentData$SampleName<-gsub("X","",as.character(treatmentData$SampleName)) #Remove the 'X' from sample names
treatmentData$SampleName<-gsub("\\..*","",as.character(treatmentData$SampleName)) #Remove all data after the '.'
treatmentData$Timepoint<-gsub("line","",as.character(treatmentData$Timepoint)) #Remove the word 'line' in the second column (Timepoint)

#rename columns of treatmentData to match with metabolomics data
colnames(treatmentData) <- c("Sample.name","Timepoint","Diet" )

# Finally, combine this data (specifically column 3 from targets data) with the metabolomics data
metabolomicsDataGroups <- merge(treatmentData, metabolomicsData,by="Sample.name")

#Sanity check: do all the timepoints (base,end) from treatmentData match with the metabolomics data time column
all(metabolomicsDataGroups$Timepoint == metabolomicsDataGroups$time) ##If this is FALSE, please check your data against the Copy you made at the start.

#Count number of columns
##columns <- ncol(metabolomicsDataGroups) ##To apply transformation on all columns
columns <- 102

##Transformation (see https://doi.org/10.1177%2F1469066720918446 and https://www.statology.org/transform-data-in-r/)
##Users can select different transformation styles here:
transformation <- "cube_root" # options are: log_2, log_10, square_root, cube_root

if(transformation == "cube_root"){
    metabolomicsDataGroups_transformed <- cbind(metabolomicsDataGroups[,c(1:4)], metabolomicsDataGroups[,5:columns]^(1/3))
}else if(transformation == "square_root"){
    metabolomicsDataGroups_transformed <- cbind(metabolomicsDataGroups[,c(1:4)], metabolomicsDataGroups[,5:columns]^(1/2))
}else if(transformation == "log_2"){
    metabolomicsDataGroups_transformed <- cbind(metabolomicsDataGroups[,c(1:4)], log2(metabolomicsDataGroups[,5:columns]))
}else if(transformation == "log_10"){
    metabolomicsDataGroups_transformed <- cbind(metabolomicsDataGroups[,c(1:4)], log10(metabolomicsDataGroups[,5:columns]))
}else{print("Warning: name for transformation not recognized")}

## Visualize the data before and after transformation (for one sample to get an idea of suitability of transformation:
#create histogram for original distribution for first column with data
hist(metabolomicsDataGroups[,5], col='steelblue', main='Original')
#create histogram for log-transformed distribution 
hist(metabolomicsDataGroups_transformed[,5], col='coral2', main=transformation)

## Test if the transformation creates a normally distributed dataset (alpha >= 0.05)
##Calculate all Shapiro values for raw and transformed data:
metabolomicsDataGroups_Shapiro <- lapply(metabolomicsDataGroups[,5:columns], shapiro.test)
metabolomicsDataGroups_transformed_Shapiro <- lapply(metabolomicsDataGroups_transformed[,5:columns], shapiro.test)

#Obtain the p-values for raw and transformed data
metabolomicsDataGroups_Shapiro_pvalues <- do.call(rbind, metabolomicsDataGroups_Shapiro)
metabolomicsDataGroups_transformed_Shapiro_pvalues <- do.call(rbind, metabolomicsDataGroups_transformed_Shapiro)

## Count how often the p-value is above 0.05, to obtain an estimate of achieved normality due to transformation
metabolomicsDataGroups_Shapiro_pvalues_sum <- sum(metabolomicsDataGroups_Shapiro_pvalues[,2] >= 0.05, na.rm=TRUE)
metabolomicsDataGroups_transformed_Shapiro_pvalues_sum <- sum(metabolomicsDataGroups_transformed_Shapiro_pvalues[,2] >= 0.05, na.rm=TRUE)

percentageThreshold <- floor(((columns)/10)*7.5) #Now set at 75% (7.5); can be strickter if the data permits this.

#Print relevant information:
if(metabolomicsDataGroups_transformed_Shapiro_pvalues_sum[1] > percentageThreshold ){paste0("Data after ", transformation ," transformation seems to follow a normal distribution for 75% or more of your data")} else{
  print("Advised to select a different data transformation procedure")}

# Visualize all information in a heatmap
heatmaply::heatmaply(metabolomicsDataGroups_transformed[,5:columns], grid_gap = 0, colors = heat.colors(200), showticklabels = c(T, F), margins = c(80, 10))
```

# Find Outliers by visualising violin plots (combination of boxplot and Kernel plot)
```{r}
# Libraries (dplyr already previously loaded but also important)
if(!"ggplot2" %in% installed.packages()){install.packages("ggplot2")}
library(ggplot2)
if(!"forcats" %in% installed.packages()){install.packages("forcats")}
library(forcats)
if(!"hrbrthemes" %in% installed.packages()){install.packages("hrbrthemes")}
library(hrbrthemes)
if(!"viridis" %in% installed.packages()){install.packages("viridis")}
library(viridis)
if(!"ggbeeswarm" %in% installed.packages()){install.packages("ggbeeswarm")}
library(ggbeeswarm)

##### Example data to check if setup of packages is correct #####
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/10_OneNumSevCatSubgroupsSevObs.csv", header=T, sep=",") %>%
  mutate(tip = round(tip/total_bill*100, 1))
  
# Grouped
data %>%
  mutate(day = fct_reorder(day, tip)) %>%
  mutate(day = factor(day, levels=c("Thur", "Fri", "Sat", "Sun"))) %>%
  ggplot(aes(fill=sex, y=tip, x=day)) + 
    geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
    scale_fill_viridis(discrete=T, name="") +
    theme_ipsum()  +
    xlab("") +
    ylab("Tip (%)") +
    ylim(0,40)
```

# Find Outliers by visualising violin plots on own data
```{r}
##### Example data to check if setup of packages is correct #####
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/10_OneNumSevCatSubgroupsSevObs.csv", header=T, sep=",") %>%
  mutate(tip = round(tip/total_bill*100, 1))

lipidClassName <- 'IDL.L'
columnNumber <- which( colnames(metabolomicsDataGroups_transformed)==lipidClassName )
maxVerticalAxis <- max(metabolomicsDataGroups_transformed[,columnNumber])
  
# Grouped
metabolomicsDataGroups_transformed %>%
  mutate(Diet =  fct_reorder(Diet, IDL.L)) %>%
  mutate(Diet = factor(Diet, levels=c("Control", "Experimental"))) %>%
  ggplot(aes(fill=Diet, y=IDL.L, x=time)) + 
    geom_violin(position="dodge", alpha=0.5, trim = FALSE, scale = "count") +
    geom_quasirandom() + ##Alternativ geom_beeswarm() + 
    scale_fill_viridis(discrete=T, name="Timepoints") +
    theme_ipsum()  +
    xlab(lipidClassName) +
    ylab("Measurement (transformed)") +
    ylim(0,maxVerticalAxis)
```


# Visualising ratio data
```{r}
ratios <- ncol(metabolomicsDataGroups)

# Visualize all information in a heatmap (remember that the ratio data was not transformed!)
heatmaply::heatmaply(metabolomicsDataGroups[,columns:ratios], grid_gap = 0, colors = heat.colors(200), showticklabels = c(T, F), margins = c(80, 10))
```


# Comparing treated to control group
```{r}
##Order columns based on Diet column (located in column names);
mSet_FINAL <- metabolomicsDataGroups_transformed[order(metabolomicsDataGroups_transformed$Diet),]

##Create one dataframe for the control group:
mSet_control <- subset(mSet_FINAL, mSet_FINAL$Diet == 'C_diet')
#Size of control group:
mSet_control_size <- nrow(mSet_control)

##Create another dataframe for the treated group:
mSet_treated <- subset(mSet_FINAL, mSet_FINAL$Diet == 'Ex_diet')
#Size of experimental group:
mSet_treated_size <- nrow(mSet_treated)

# The groups are of unequal size (which you would normally want to compensate for; we will take the data as it is)
##The unpaired two-samples t-test is used to compare the mean of two independent groups.
# T-test for one specified column:
#tTestresults <- t.test(mSet_control$XXL.VLDL.P, mSet_treated$XXL.VLDL.P, alternative = "two.sided", var.equal = FALSE)

# Ttest for all columns:
grp <-mSet_FINAL$Diet
tests_list <- lapply(names(mSet_FINAL[5:ncol(mSet_FINAL)]), function(x) t.test(as.formula(paste0(x, "~ grp")), data = mSet_FINAL[,5:ncol(mSet_FINAL)]))

result <- do.call(rbind, lapply(tests_list, `[[`, "estimate"))
pval <- sapply(tests_list, `[[`, "p.value")
result <- cbind(names(mSet_FINAL[5:ncol(mSet_FINAL)]), result, p.value = pval)

```


# Use Bubble chart to inspect cleaned up data from Nightingale
```{r}
if(!"bubbleHeatmap" %in% installed.packages()){install.packages("bubbleHeatmap")}
library(bubbleHeatmap)

##### Example data to check if setup of packages is correct #####

#> Loading required package: grid
#> 
myData <- merge_template(cetp, "ckb_id")
metabTree <- metabFigure(myData)

gridData <- formatData(myData, colorValue="estimate", sizeValue = "negLog10P", nightingale = TRUE)

treeInput <- multiPlotInput(colorList=gridData$colorList, 
                            sizeList=gridData$sizeList, 
                            nightingale=TRUE, legendHeight=8)

treeList <- bubbleHeatmapList(treeInput)

metabTree <- metabFigurePlot(treeList)

grid.newpage()
grid.draw(metabTree)
```


# Use Bubble chart to inspect cleaned up data from Nightingale on own data TODO
```{r}
if(!"bubbleHeatmap" %in% installed.packages()){install.packages("bubbleHeatmap")}
library(bubbleHeatmap)

##### Example data to check if setup of packages is correct #####

#myData <- merge_template(cetp, "ckb_id")
#metabTree <- metabFigure(metabolomicsDataGroups_transformed)

#gridData <- formatData(myData, colorValue="estimate", sizeValue = "negLog10P", nightingale = TRUE)

#treeInput <- multiPlotInput(colorList=gridData$colorList, 
#                            sizeList=gridData$sizeList, 
#                            nightingale=TRUE, legendHeight=8)

#treeList <- bubbleHeatmapList(treeInput)

#metabTree <- metabFigurePlot(treeList)

#grid.newpage()
#grid.draw(metabTree)
```